% --- Author Bio ---
\startcentered
\frontmattergap
Paul M. Cray was born in Preston, Lancashire in 1968.
He has an MA in Physics from Oxford,
a PhD in neutron scattering from Salford and an MBA from Imperial.
He lives with his wife in Seattle, Washington.
{\twelvesl THE VISIBILITIES} is his magnum opus, a recursive exploration of the seen and unseen.
\stopcentered

\blankpage

% --- Half Title ---
\startcentered
\frontmattergap
{\twelvess THE VISIBILITIES}
\stopcentered

\blankpage

% --- Full Title Page ---
\startcentered
\frontmattergap
{\seventeenss THE VISIBILITIES\par}
\smallskip
{\twelvesl A Magnum Opus\par}
\titlegap
{\seventeenss PAUL M.~CRAY\par}
\publishergap
{\twelvebf ATOMIC RAZOR PRESS}
{\twelvebf Seattle and Preston}
\stopcentered

% --- Copyright Page ---
\startcentered
\twelverm
PUBLISHED BY ATOMIC RAZOR PRESS
\copyrightgap
Copyright \copyright\ Paul M.~Cray 2026
\copyrightgap
Paul M.~Cray has asserted his right under the Copyright, Design
and Patents Act 1988 to be identified as the author of this work.
\copyrightgap
{\twelveit This book is sold subject to the condition that it shall not, by way of}
{\twelveit trade or otherwise, be lent, resold, hired out, or otherwise circulated}
{\twelveit without the publisher's prior consent in any form of binding or cover}
{\twelveit other than that in which it is published and without a similar condition}
{\twelveit including this condition being imposed on the subsequent publisher.}
\copyrightgap
First published in the United States in 2026 by
Atomic Razor Press
\copyrightgap
Typeset with \OpTeX\ by Atomic Razor Press
\stopcentered

% --- Dedication ---
\startcentered
\frontmattergap
{\twelveit For those who look closely enough to see the gaps.}
\stopcentered

\blankpage

% --- Global Epigraphs ---
\null\vfill
\begingroup
\leftskip=35mm \rightskip=20mm
\parindent=0pt

\it ‘The past is never dead. It’s not even past.’
\smallskip
\rm \noindent --- William Faulkner, {\it Requiem for a Nun}

\bigskip
\it ‘The past is a foreign country; they do things differently there.’
\smallskip
\rm \noindent --- L.P. Hartley, {\it The Go-Between}

\bigskip
\it ‘Why do we remember the past but not the future?’
\smallskip
\rm \noindent --- Stephen W. Hawking, {\it A Brief History of Time}

\bigskip
\it ‘History is a nightmare from which I am trying to awake.’
\smallskip
\rm \noindent --- James Joyce, {\it Ulysses}
\endgroup
\vfill\eject

\blankpage
