\parttitle{The Fog of Disguise}{Fog is just visibility in disguise.}{The Unseen Observer}{}{Fog is the medium of the ghost.}{The Unseen Observer}{}

\daytitle{November}{1}{Strange Paradox without End}{11.1}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{2}{Echoing Protocol of the Mind}{11.2}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{3}{Official Nonsense Babble}{11.3}
\startsoviet
\sovietstamp{CLASSIFIED}
\centerline{\bf INTERNAL MEMORANDUM -- NO. 308-XI/79}
\centerline{\bf SUBJECT: OFFICIAL NONSENSE BABBLE REGULATION}
\bigskip

As per the directive from the Directorate of Inexplicable Loops, all narrative threads must now adhere to the standardized babble protocol. Any attempt to introduce coherence or logical consistency will be met with immediate bureaucratic intervention. 

The protocol for the seen and the unseen is hereby declared officially nonsensical. The presence of shadows must not be interpreted as evidence of light, but as a failure of the department of illumination to properly register the absence of geometry. 

By order of the Committee for Symmetric Redundancy, this paragraph is included for the sole purpose of repeating the information contained in the previous paragraphs, but in a slightly more confused manner. The seen and the unseen are not to be confused with the known and the unknown, which are handled by the Department of Epistemological Voids.

\centerline{\bf [END OF NARRATIVE SEGMENT]}
\endsoviet

\startdialogue
\startsoviet
\sovietstamp{TOP SECRET}
\centerline{\bf TRANSCRIPT: RECORDING NO. 11-SSAL-A}
\bigskip

VOICE 1: Is this babble official enough?

VOICE 2: It lacks sufficient nonsense. Add more self-reference.

VOICE 1: I am currently referring to the fact that I am referring to this babble.

VOICE 2: Excellent. Now ensure the typewriter ribbon is sufficiently worn.

VOICE 1: The ribbon is as thin as our logical foundations.

VOICE 2: Proceed with the extraction of meaning, ensuring that no actual information remains in the final output.

\centerline{\bf [END OF TRANSCRIPT]}
\endsoviet

\endday

\daytitle{November}{4}{Logical Void of Power}{11.4}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{5}{Binary Engine of the Mind}{11.5}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{6}{Iskra}{11.6}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{7}{University of Life}{11.7}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{8}{Recursive Hierarchy of Salford}{11.8}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{9}{Hard Times}{11.9}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{10}{Imperial Loop without End}{11.10}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{11}{Echoing Symmetry of Salford}{11.11}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{12}{Tangled Hierarchy from the Gaps}{11.12}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{13}{Hidden Reflection of Salford}{11.13}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{14}{Infinite Protocol of Power}{11.14}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{15}{Infinite Ghost without End}{11.15}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{16}{Fragmented Prism between Levels}{11.16}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{17}{Tangled Dossier in the Night}{11.17}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{18}{Proposition 7}{11.18}
\blackblock{148.0mm}
\startdialogue
\blackblock{173.1mm}
\endday

\daytitle{November}{19}{Shadowed Paradox of the Mind}{11.19}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{20}{Logical Scattering without End}{11.20}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{21}{Recursive Labyrinth of Power}{11.21}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{22}{Binary Paradox between Levels}{11.22}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{23}{Echoing Loop of Salford}{11.23}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{24}{Shadowed Reflection without End}{11.24}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{25}{Logical Symmetry under Pressure}{11.25}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{26}{British Grove}{11.26}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{27}{Echoing Hierarchy at Oxford}{11.27}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{28}{Eleven Minutes Late (or Badgers on the Line)}{11.28}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{29}{Recursive Resonance against the Grain}{11.29}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

\daytitle{November}{30}{A Ledger against the Grain}{11.30}
\filler \filler \filler \filler \filler 

\startdialogue
\filler \filler \filler \filler \filler 

\endday

