\chaptertitle{Credits}

                                       Credits
Figures: Fig. 1, Johann Sebastian Bach by Elias Gottlieb-Hart ssmann (1748), collection
of William H. Scheide, Princeton, New Jersey; Fig. 2, Flute Concert in Saussouci, by
Adolf von Menzel, Nationalgalerie, West Berlin; Figs. 3, 4, 152, "The Royal Theme" and
the last page of the "Six-part Ricercar." from the original edition of Musical Offering by
Johann Sebastian Bath, are reproduced courtesy of the Library of Congress; Figures of
lithographs and woodcuts of M. C. Escher are reproduced by permission of the Escher
Foundation, Haags Gemeentemuseum, The Hague, copyright © the Escher Foundation,
1979, reproduction rights arranged courtesy of the Vorpal Galleries, New York, Chicago,
San Francisco, and Laguna Beach; Fig. 9, photograph of Kurt Godel by Orren J. Turner
from Foundations of
Mathematics. Symposium Papers Commemorating the Sixtieth Birthday of Kurt Godel,
edited by
Jack J. Bulloff, Thomas C. Holyoke, and S. W. Hahn, New York, Springer-Verlag, 1969;
Figs. 17, 96, "Figure-Figure" and "A section of mRNA passing through a ribosome,"
drawings by Scott E. Kim; Figs. 19, 44, 133, 148, musical selections from the Musical
Offering by J. S. Bach, music printed by Donald Byrd's program "SMUT"; Fig. 25,
"Cretan labyrinth" from W. H. Matthews, Mazes and Labyrinths: Their History and
Development, New York, Dover Publications, Inc., 1970; Fig. 39, photograph of Rosetta
Stone, courtesy of the British Museum; Fig. 40, A collage of scripts. Samples of
cuneiform, Easter Island, Mongolian and Runic scripts from Hans Jensen, Sign, Symbol
and Script, East Germany VEB Deutscher Verlag Der Wissenschaften; samples of
Bengali and Buginese script from Kenneth Katzner, The Languages of the World, New
York, Funk \& Wagnalls, 1975; samples of Tamil and Thai from I. A. Richards and
Christine Gibson, English Through Pictures, New York, Washington Square Press; Fig.
59, "Intelligence built up layer by layer" (Fig. 9.8) adapted from Patrick Henry Winston.
Artificial Intelligence, Reading, Mass., Addison-Wesley Publishing Company, reprinted
by permission; Figs. 63, 69, photographs of an ant bridge by Carl W. Rettenmeyer and
construction of an arch by termite workers by Turid Hdlldobler, from E. O. Wilson, The
Insect Societies, Cambridge, Mass., Harvard University Press, 1979; Fig. 65, schematic
drawing of a neuron adapted from The Machinery of the Brain, by Dean Wooldridge,
copyright © 1963, McGraw-Hill, Inc. used with permission of McGraw Hill Book
Company, and from Fig. 11-6, page 26, Speech and Brain-Mechanisms, by Wilber
Penfield and Lamar Roberts, copyright © by Princeton University Press, reprinted by
permission of Princeton University Press; Fig. 66, "the human brain, seen from the left
side," from Steven Rose, The Conscious Brain, copyright © 1973 by Steven Rose,
reprinted by permission of Alfred A. Knopf, Inc., New York, and John Wolfers, London;
Fig. 68, "overlapping neural pathways," from John C. Eccles, Facing Reality, New York,
SpringerVerlag, 1970; Figs. 77, 78, 80, 82, 117, 137, 138, 141, The Shadows, State of
Grace, The Fair
Captive, The Air and the Song, Mental Arithmetic, Common Sense, The Two Mysteries,
and The
Human Condition I by Rene Magritte, copyright © by ADAGP, Paris, 1979; Figs. 79, 95,
"Tobacco Mosaic Virus" and "Secondary and Tertiary Structure of Myoglobin" from
Albert Lehninger, Biochemistry, New York, Worth Publishers, 1975; Figs. 91, 92, "The

Credits                                                                               759

four constituent bases of DNA" and "The ladder-like structure of DNA" from Arthur
Kornberg, "The Synthesis of DNA," Scientific American, copyright © October 1968, all
rights reserved; Fig. 93, "Molecular model of the DNA double helix," reprinted by
permission, from V. M. Ingram, Biosynthesis of Macromolecules, Menlo Park,
California, The Benjamin/Cummings Publishing Company, 1972; Fig. 97,
"Polyribosome" from The Proteins, edited by R. E. Dickerson and H. Neurath, page 64,
New York, Academic Press; Fig. 98, "A two-tiered molecular canon" from O. L. Miller,
Jr., "Visualization of Genes in Action," in Scientific American, copyright © March 1973,
all rights reserved; Figs. 101, 102, 103, "The T4 bacterial virus," "Infection of a
Bacterium by a virus," and "The morphogenetic pathway of the T4 virus," from William
B. Wood and R. S. Edgar, "Building a Bacterial Virus" in Scientific American, copyright
© July 1987, all rights reserved; Fig.
105, photograph of Srinivasa Ramanujan from S. R. Ranganathan, Ramanujan, the Man
and the Mathematician, New York, Asia Publishing House, 1967; Figs. 110, 111, 112,
from Terry Winograd, Understanding Natural Language,New York, Academic Press,
1972; Fig. 113, Photograph of Alan Turing by Mssrs. C. H. O. Trevelyan from Sara
Turing, Alan M. Turing, Cambridge, England, W. H. Heffer and Sons, Ltd., 1959; Fig.
116, "a meaningful story in Arabic" from Abdelkebir Khatibi and Mohammed Sijelmassi,
The Splendor of Islamic Callizrabhl, New York, London, Thames \& Hudson, copyright
© by Qarawiyne Library in Fez. Fig 118, Procedural representation of “red cube which
supports a pyramid,” adapted from Computer Models of Thought and Language, edited
by Roger C. Schank and Kenneth Mark Colby. W. H. Freeman and Company, copyright
1973; Figs. 119, 122, 124, 130, Bongard problems from M. Bongard, Pattern
Recognition, Rochelle Park, New Jersey,
Hayden Book Company, Spartan Books, 1970

Grateful acknowledgment is made to the following publishers for permission to quote
excerpts from the following material: The Bach Reader. A Life of johann Sebastian Bach
in Letters and Documents, edited by Hans T. David and Arthur Mendel, Revised, with the
permission of W. W. Norton \& Company, Inc., copyright (~) 1966, 1945 by W. W.
Norton \& Company, Inc. copyright renewed 1972 by Mrs. Hans T. David and Arthur
Mendel; J. S. Bach's Musical Offering, page 179, edited by Hans T. David, New York,
copyright 1945 by G. Schirmer, Inc., used by permission; Gyomay Kubose, Zen Koans,
Chicago, Regnery, 1973; Pauls Reps, Zen Flesh, Zen Bones, Tokyo, Japan, Charles E.
Tuttle Co., Inc. 1957; J. R. Lucas, "Minds, Machines, and Godel," and Alan M. Turing,
"Computing Machinery and Intelligence," from ,finds and Machines, edited by A. R.
Anderson, Englewood Cliffs, New Jersey, Prentice-Hall, 1964, and Philosophy, vol. 36,
1961; J. M. Jauch, Are Quanta Real? Bloomington, Indiana, Indiana University Press,
1973; James R. Newman, "Srinivasa Ramanujan," in The World of Mathematics, edited
by James R. Newman, New York, Simon \& Schuster, reprinted by permission of Simon
\& Schuster, a division of Gulf \& Western Corporation, 1956; Terry Winograd, "A
Procedural Model of Language Understanding," from Computer Models of Thought and
Language, edited by Roger C. Schank and Kenneth Mark Colby, San Francisco, W. H.
Freeman and Company, copyright © 1973; Joseph Weizenbaum, Computer Power and
Human Reason: From judgment To Calculation, San Francisco, W. H. Freeman and
Company, copyright © 1976.

Credits                                                                             760

