\chaptertitle{Introduction: A Musico-Logical Offering}

                                                Notes
                               Introduction: A Musico-Logical Offering

1 H T David and A. Mendel, The Bach Reader,               Memoir "Sketch of the Analytical Engine
pp. 305-6.                                                Indented by Charles Babbage", by 1.. F.
2 Ibid., p. 179                                           Menabrea (Geneva, 1842), reprinted in P.
3 Ibid., p. 260                                           and E. Morrison, Charles Babbage and His
4 Charles Babbage. Passages from the Li/e of              Calculating Engines, pp. 248-9, 284.
Philosopher, pp. 145-6.                                   6 David and Mendel, pp. 255-6. ¨
5 Lady A. A. Lovelace, Notes upon the                     7 Ibid.. p. 40.

                                            Two-Part Intention

1 Lewis Carroll, "What the Tortoise Said to Achilles", .Wind, ns., 4 (1895), pp. 278-80.

                        Chapter IV: Consistency, Completeness, and Geometry

1 Herbert Meschkovcski. Non-Euclidean                     2 Ibid., p. 33.
Geometry. pp. 31-2

                                 Chapter VI: The Location of Meaning

1 George Steiner..After Babel, pp. 172-3.                 .2 Leonard B. Meter. Music, The Arts, and Ideas,
                                                          pp. 87-8

                                Chapter VII: The Propositional Calculus

1 Gyomay M. Kubose, Zen Koans, p. 178                     3 A. R. Anderson and N. D. Belnap, Jr. .
2 Ibid., p. 178.                                          Entailment (Princeton, N.J.: Princeton University
                                                          Press. 1975).

                                              A Mu Offering

1 All genuine koans in this Dialogue are taken from Paul Reps, Zen Flesh, Zen Bones, Gvomas M. Kubose,
Zen Koans

                                    Chapter IX: Mumon and Godel

1 Paul Reps, Zen Flesh, Zen Bones, pp. 110-11             7 Reps, p. 121.
2 Ibid., p. 119.                                          8 Gyomay M. Kubose Zen Koans p. 35.
3 Ibid., pp. 111-12.                                      9 Zen Buddhism, p. 31.
4 Zen Buddhism (Mount Vernon, N.Y.: Peter                 10 Kubose, p. I10
Pauper Press, 1959). p. 22                                11 Ibid., p. 120.
5 Reps, p. 124.                                           12 Ibid., p. 180.
6 'Zen Buddhism, p. 38.                                   13 Reps, pp. 89-90.

                                   Chapter XI: Brains and Thoughts

1 Carl Sagan, ed. Communication with                      , pp 251-2.
Extraterrestrial Intelligence, p. 78.                     3 E. O. Wilson, The Insect Societies, p. 226.
2 Steven Rose, The Conscious Brain                        4 Dean Wooldridge,:Mechanical Man, p. 70.




Notes                                                                                                 738

                                      English French German Suite

1 Lewis Carroll. The Annotated Alice (Alice's      versions. The original sources for the French and
Adventures in Wonderland and Through the           German texts are given below.
Looking-Glass). Introduction and Notes by Martin   2 Frank L. Warrin, The Vew Yorker, Jan. 10, 1931.
Gardner (New fork: Meridian Press, New American 3 Robert Scott, "The Jabberwock Traced to Its True
Library, 1960). This source contains all three     Source", Macmillan's Magazine, Feb. 1872
.
                                   Chapter XII: Minds and Thoughts

1 Warren Weaver, "Translation     in Machine          2 C. H. MacGillavrv. Symmetry Aspects of the
Translation of Languages, Wm. N. Locke and A.         Periodic Drawings of M C. Escher, p. viii,
Donald Booth. eds. (News fork: John Wiles and         3 J. R. Lucas, "Minds, Machines, and Gödel", in A.
Sons, and Cambridge, Mass,: SLI. F. Press, 1955),     R. Anderson, ed., Minds and Machines, pp. 57-9.
p. 18.

                              Chapter XIII: BlooP and FlooP and GlooP

1 J . M. Jauch, Are Quanta Real?, pp. 63-65.

                        Chapter XIV: On Formally Undecidable Propositions
                                   of TNT and Related Systems

1 The title of Gödel’s 1931 article included a Roman the difficult arguments. However, the first paper was
numeral "I" at the end, signifying that he intended to so widely acclaimed that a second one seas rendered
follow it up with a more detailed defense of some of superfluous, and it was never written

                               Chapter XV: Jumping out of the System

1 Lucas in Anderson, p. 43.                           4 M. C. Escher, The Graphic Wash of M. C Escher
2 Ibid., p. 48.                                       (New York: Meredith Press. 1967), p. 21.
3 Ibid., pp. 48-9.                                    5 Ibid., p. 22.
                                                      6 E. Goffinan. Frame Analysis, p. 475.

                                 Edifying Thoughts of a Tobacco Smoker

This translation of Bach's poem is taken from Day id and Mendel, The Bach Reader, pp. 97-8.

                         Chapter XVII: Church, Turing, Tarski, and Others

1 Stanislaw Ulam, Adventures of a Mathematician,      5 Newman, p. 375.
p. 13.                                                6 Ibid., p. 375.
2 James R. Newman, "Srinivasa Ramanujan°, in          7 Ibid., p. 375-6.
James R. Newman, ed., The World of Mathematics        8 Ibid., p. 376.
(New York: Simon and Schuster, 1956), Vol. 1, pp.     9 Lucas in Anderson, p. 44.
372-3.                                                10 Ibid., p. 54.
3 Ibid., p. 375.                                      11 Ibid., p. 53.
4 S. R. Ranganathan, Ramanujon, pp. 81-2.


                                   SHRDLU, Toy of Alan's Designing

1 This Dialogue is adapted front 'terry Winograd, "A Thought and Language, pp. 155-66. Only the names
Procedural Model of Language Understanding", in      of two characters have been modified.
R. Schank and K.Colby, eds., Computer Models of




Notes                                                                                                744

