\chapter{XVIII}
\chaptertitle{Artificial Intelligence: Retrospects}

                          Chapter XVIII: Artificial Intelligence: Retrospects

1 Alan M. Turing, "Computing Machinery and             and J. W. Beauchamp, eds., Music by Computers, p.
Intelligence", Mind, Vol. LIX, No. 236                 96.
(1950). Reprinted in A. R. Anderson, ed., Minds and    13 Ibid., p. 106.
Machines.                                              14 Carl Sagan, Communication with Extraterrestrial
2 Turing in Anderson, p. 5. 3 Ibid, p. 6               Intelligence, p. 52.
4 Ibid., p. 6.                                         15 Art-Language, Vol. 3. No. 2, May 1975.
5 Ibid., P. 6.                                         16 Terry Winograd, "A Procedural Model of
6 Ibid., pp. 13-4.                                     Language Understanding", in R. Schank and K.
7 Ibid., pp, 14-24.                                    Colbc, eds.. Computer Models of Thought and
8 Ibid., p. 17                                         Language, p. 170.
9 Vinton Cerf, "Parry Encounters the Doctor", p. 63.   17' Ibid., p. 175.
10 Joseph Weizenhaum, Computer Power and               18 Ibid..- p. 175.
Human Reason, p. 189.                                  19 Terry Winograd, Understanding Natural
11 Ibid., pp. 9-10                                     Language, p. 69.
12 M. Mathews and L. Rosier, "A Graphical              20 Winograd, "A Procedural Model", pp. 182-3.
Language for Computer Sounds" in H. vn Foerster        21 Ibid., pp. 171-2.


                            Chapter XIX: Artificial Intelligence: Prospects

1 The New Yorker, Sept. 19, 1977, p. 107.
2 Ibid., p. 140.
3 George Steiner, After Babl(, pp. 215-227.
4 David E. Rumelhart, "Notes on a Schema for Stories", in D. Bobrow and A. Collins, eds., Representation
and Understanding, p. 211.
5 Stanislaw Ulam, Adventures of a Mathematician, p. 183.
6 Marvin Minsky, "Steps Toward Artificial Intelligence", in E. Feigenbaum and J. Feldman, eds.,
Computers and Thought, p. 447.
7 Ibid., p. 446.

                        Chapter XX: Strange Loops, Or Tangled Hierarchies

1 A. L. Samuel, "Some Moral and Technical Consequences of Automation-A Refutation", Science 132
(Sept. 16, 1960), pp. 741-2.
2 Leonard B. Meyer, Music, The Arts, and Ideas, pp. 161, 167.
3 Suzi Gablik, Magritte, p. 97.
4 Roger Sperry, "Mind, Brain, and Humanist Values", pp. 78-83.
5 H T. David, J. S. Bach's Musical Offering, p. 43.




Notes                                                                                               744

